\section{RSA}

Now for RSA $\ldots$

We only need to work with integers. Why? Because any message $M$ is
really just a sequence of bits and you can cut your sequences of
bits which can be viewed as an unsigned int, i.e, an integer $\geq 0$.
So we'll just think of our messages as integers.

So once again suppose Alice wants to send a secret to Bob. The
secret is an integer $x$.

Bob selects two distinct primes $p$ and $q$.
He computes $N = pq$ and selected two positive integers $e$ and $d$
such that
\[
 ed \equiv 1 \,\,\,(\operatorname{mod} \phi(N))
\]
In other words $e$ and $d$ and multiplicative inverses of each
other in $\Z/\phi(N)$.
Furthermore Bob publishes $N$ and $e$ publicly.
$e$ is called the \defone{encryption exponent}
and
$d$ is called the \defone{decryption exponent}.

We assume that the secret $x$ is less than $N$
because we'll be working in $\Z/N$.
(Again if $x$ as a bit sequence is too large for $N$,
then we cut $x$ up into smaller block of bits and send them
separately.)

Since $N$ and $e$ are public, Alice can download $N$ and $e$
and then compute
\[
E_{(N,e)}(x) = x^e \,\,\,(\operatorname{mod} N)
\]
I'll write $x^e \,\,\,(\operatorname{mod} N)$ for the 
least positive remainder of $x^e$ mod $N$.
She then sends $x^e \,\,\,(\operatorname{mod} N)$ to Bob.

When Bob received $x^e \,\,\,(\operatorname{mod} N)$, he computes
\[
D_{(N,d)}(x) = x^{ed} \,\,\,(\operatorname{mod} N)
\]
What we need to prove is that RSA works, i.e.,


\begin{thm}
  Let $p,q$ be primes and $N = pq$.
  If $ed \equiv 1 \,\,\,(\operatorname{mod} \phi(N))$,
  then
  \[
  (x^e)^d \equiv x \,\,\,(\operatorname{mod} N)
  \]
  for all integers $x$.
\end{thm}

\proof
TODO
\qed

\begin{prop}
  Prove the following:
  \begin{enumerate}[nosep,label=\textnormal{(\alph*)}]
  \item
    If $p$ and $q$ are distinct primes such that $p \mid a$ and $q \mid a$,
    then $pq \mid a$.
  \item 
    If $x \mid a$ and $y \mid a$ and $\gcd(x,y) = 1$, then $xy \mid a$.
  \item
    If $x \mid a$ and $y \mid a$, then $(xy/\gcd(x,y)) \mid a$.
  \end{enumerate}
  (b) is a generalization of (a)
  and (c) is a generalization of (b).
  \qed
\end{prop}
\proof
(a)
TODO

(b)
TODO

(c)
TODO
\qed

Note that RSA (and all public key ciphers) are not
meant for encrypting messages like strings (emails, sales receipts,
image files, etc.)
This is an unfortunate thing that's done in many cryptography textbooks. 
They are actually used to encrypt/decrypt keys for private ciphers
since private ciphers are much faster.

For instance suppose Bob wants to send an encrypted message $M$ to Alice.
He first picks a private cipher.
Say, he is going to use the substitution cipher.
He creates a key for the substitution cipher:
$a \mapsto h, b \mapsto z, c \mapsto t, ...$.
The key can be  described by the string $\texttt{key} = hzt...$.
This is of course a binary sequence of $26 \times 8 = 208$ bits.
This can of course be viewed as a positive integer mod $N$ as long as
$N$ is large enough.
Bob can use RSA to send the encrypted \verb!key! to Alice,
which Alice will decrypt.
After that Bob can send his $M$, encrypted using the substitution cipher
with \verb!key! to Alice.

Therefore RSA (and other public key cipher) is used to
setup keys for private key cipher (example: AES) because
AES is much faster than RSA and the message to be sent is usually way
longer than the key.

The speed of private ciphers
should be clear from the chapter on classical ciphers.
Likewise modern private ciphers such as 3DES, AES, etc. are much faster
than RSA and other private key ciphers.
The current modern private cipher is AES.
3DES is considered deprecated since 2018.
In particular the RSA standard
(called PKCS -- you can easily find lots of webpages on PKCS)
does not include
specification on how to break up long messages before encryption and how to
reassembly them after decryption.

OK, let's summarize everything.
In the following, I'll write $x \pmod{N}$ to be the remainder when $x$
is divided by $N$.
There are three steps: Bob has to
generate keys, Alice has to encrypt, and Bob has to decrypt.
\begin{enumerate}[nosep]
 \item Key Generation:
  \begin{enumerate}
   \item Bob selects distinct primes $p$ and $q$.
   \item Bob computes $N = pq$.
   \item Bob computes $\phi(N) = (p-1)(q-1)$.
   \item Bob selects $e$ such that $0 < e < \phi(N)$,
   $\gcd(e, \phi(n)) = 1$.
   \item Bob computes $d$ such that
     $ed \equiv 1 \pmod{\phi(N)}$.
 \item Bob publishes $(N,e)$ (the public key) but keeps $(N,d)$
   (the private key) to himself.
  \end{enumerate}
 \item Encryption:
   Alice obtains the publicly available $(N,e)$ (the public key) and
   computes
  \[
   E_{(N,e)}(x) = x^e \,\,\,(\operatorname{mod} N)
  \]
  and sends it to Bob.
 \item Decryption: Bob uses $(N,d)$ (the private key) to compute
 \[
  D_{(N,d)}( x^e \,\,\,(\operatorname{mod} N) ) = x^{ed} \,\,\,(\operatorname{mod} N)
 \]
\end{enumerate}

Note that the key is made up of
\begin{enumerate}[nosep]
  \li a \defone{public key} $(N,e)$ and
  \li a \defone{private key} $(N,d)$
\end{enumerate}
$(N,e)$ is revealed to the public.
$(N,d)$ is kept private.
In general

\begin{defn}
  A \defone{public cipher} is made up of the encryption
  and decryption functions
  $E_{\text{pubkey}}, D_{\text{privkey}}$ which depends on
  the key
  $k = (\text{pubkey}, \text{privkey})$
  which is a 2-tuple made up of
  the public key and private key.
  Such a cipher is also called an
  \defone{asymmetric key cipher}.
\end{defn}

Recall that in the case of private (or symmetric) key cipher
the encryption and decryption keys are the same. 

\input{exercises/rsa-14/main.tex}
