\sectionthree{Modular arithmetic}
\begin{python0}
from solutions import *; clear()
\end{python0}

Number theory is basically the study of whole numbers.
That however is an over-simplification.
The study of number theory involves almost all areas of mathematics.
In fact many areas of mathematics were created just to study 
certain problems in number theory.

Number theory is an extremely huge area of study in Mathematics and Computer
Science. It is also extremely fascinating. Many problems is number theory can
be stated very simply so that even a high school student can understand the
statement of the problem. And yet the techniques used to solve some of these
problems require the mathematical tools from almost every area of Math. Gauss
once said that ``Mathematics is the queen of sciences and number theory is the
queen of mathematics".

Although there are many branches within Number Theory, right now
we only need to know a little bit about Elementary Number Theory.
``Elementary" here does not mean simple (although it will be for
us since you're only seeing a small part of Elementary Number
Theory). It means we are studying Number Theory using only
properties of whole numbers (integers). Research in Number Theory
requires real numbers, complex numbers, calculus, geometry,
complex analysis, etc.

This will be a very short introduction to the vast area of Number
Theory. In fact this is only a tiny fraction of Elementary Number
Theory. This is one of the oldest area of Mathematics and one of
the fascinating because of its history. If you want to learn more
about number theory, just let me know. I can easily find a project
for you to work on.

For now, we will look at modular arithmetic.
Besides cryptography,
modular arithmetic is also used in
data compression and error correction codes.

The set of integers $\{..., -3, -2, -1, 0, 1, 2, 3, ...\}$
denoted $\Z$ has two operations $+$ and $\cdot$.
In terms of the algebraic structure (i.e. the operations), $\Z$ is known as
a \textbf{commutative ring}.
Basically a commutative ring is a set of \lq\lq things'' with
two operations, addition and multiplication, with rules that
look like the addition and multiplication rules for $\Z$.
For instance one such rule in $\Z$ is
\[
  a (b + c) = ab + ac
\]
This same rule holds true for $\Q$, $\R$, $\C$ and polynomials with coefficients in $\Z$.

The reason why mathematicians even bother defining this concept
of \lq\lq commutative ring"
is more or less
the same reason why we write functions in programs: for re-use.
There are \textit{many} naturally occurring rings.
So if while developing the theory for 
$\token{\operatorname{blah_1}}$
and
$\token{\operatorname{blah_2}}$
and they are both rings, then it's enough to prove a general fact that applies
to both and quote the fact.
This is also related to the concept of inheritance and abstract base classes.
You can think of $\Z, \Q, \R, \C$ as subclasses of \verb!CommutativeRing!.
Therefore if you have a function
\begin{console}
void f(CommutativeRing & r)
{
  ...
}
\end{console}
then \verb!f! can work with \verb!x! if \verb!x! is a object of $\Z, \Q, \R, \C$.

The mathematician develop  general results while the programmer
write general functions working on abstract base class objects.
The idea is the same.
The reason for generality is efficiency.

Here's a very important advice on studying rings, groups, fields, math, etc.
(in fact this applies to any area of study where there is a great deal of 
generalization):
Although the definition and theorems are general, you
\textit{always} keep a
couple of standard examples in your mind as you read the statements.
While reading them, mentally substitute your examples into the facts so that 
you can associate it to something more concrete.
This is not just a learn technique for undergraduate students.
Even researchers do that when they read research papers.
For the definition of ring below, think of the ring of integers.

I will try to be as informal as possible for now
so that you can develop some feel/intuition for what we need for now.
Later I'm going to come back to this topic and redo the whole thing rigorously.
The focus for now is to understand modulo 26 arithmetic.

Anyway, a \textbf{commutatative ring} $R$ 
\sidenote{
Example: the integer $\Z$,...$+$ and $\cdot$ of $\Z$ ...
}
is a set of \lq\lq things''
with two operations abstractly denoted by 
$\oplus$ and $\odot$ (\lq\lq addition'' and \lq\lq multiplication'')
Furthermore there are two special \lq\lq things'' in $R$ which we will call
$0_R$ and $1_R$.
The properties satisfied by $R,0_R,1_R,\oplus,\odot$ 
\sidebar{
0 and 1 of $\Z$ ...
}
(of course there must be
something satisfied by them!) are as follows.
For $\oplus$ the properties are:
\begin{myenum}
\item $r \oplus s$ is in $R$ for all $r,s$ in $R$
\sidebar{$i + j$ is in $\Z$ for integers $i$, $j$, ...}
\item 
$(r \oplus s) \oplus t
=
r \oplus (s \oplus t)$
for all $r,s,t$ in $R$
\sidebar{
$(i + j) + k = i + (j + k)$
for integers $i,j,k$ ...
}
\item
$r \oplus 0_R = r = 0_R \oplus r$ 
for all
$r$ in $R$
\sidebar{
$i + 0 = i = 0 + i$
for integer $i$ ...
}
\item For 
\sidebar{
If $i$ is an integer, then $-i$ is also an integer and 
$i + (-i)= 0 =$
$(-i)+ i$.
}
all $r$ in $R$ there is something in $R$ which we will call $r'$
such that $r \oplus r' =$ $0_R = r' \oplus r$
\end{myenum}
For $\odot$ the properties are:
\begin{myenum}
\item $r \odot s$ is in $R$ for all $r, s$ in $R$
\sidebar{$ij = ji$ for integer $i,j$,...}
\item $(r \odot s) \odot t = r \odot (s \odot t)$ for all $r,s,t$ in $R$
\sidebar{
$(ij)k = i(jk)$ for integers $i,j,k$
}
\item $r \odot 1_R = r = 1_R \odot r$ for all $r$ in $R$
\sidebar{
$i1 = i = 1i$ for integer $i$ ...
}
\item $r \odot s = s \odot r$ for all $r,s$ in $R$ 
\sidebar{
$ij = ji$ for integers $i,j$
}
\end{myenum}
The property involving both $\oplus$ and $\odot$ is
\begin{myenum}
\item 
$r \odot (s \oplus t) =$
$r \odot s \oplus r\odot t$
\sidebar{
$i(j + k) = ij + ik$ for integers $i,j,k$.
Phew! So $\Z$ is a commutative ring.
}
\end{myenum}

Just remember this: 
A ring is a set of things with addition and multiplication.
And when you're lost just think of the set of integers and its operations.

Now we'll be working with the alphabet a,b,...,z.
We'll call them by their new identities: $0,1,...,25$.
This is not exactly all of $\Z$.
The formulas for the encryption and decryption of Caesar's cipher
involves addition and
substraction.
What if we go beyond? 
What is 3 + 25 (i.e., d + z)?
No problem, we will take remainders mod 26.
\sidebar{In C\texttt{++}-speak, we take $\% 26$.}
So instead of 3 + 25 we think of (3 + 25) mod 26 instead.
Of course the remainders are 0, 1, ..., 25 which is exactly what we want. 

To indicate that we're only interested in remainders or more accurately, we ignore multiples of 26, we write
\begin{align*}
3 + 25 
&= 28 \\
&\equiv 2 \pmod{26}
\end{align*}
In general, if $x$ and $y$ are integers, we write
\[
x \equiv y \pmod{26}
\]
if 26 divides $x - y$.
It does not mean that $x$ is equal to $y$.
It means that $x$ and $y$ are the same if you ignore additive multiples of 26, i.e.,
\[
x \equiv y \pmod{26}
\]
is the same as saying
\[
  x = y + (\text{... some multiple of 26 ...})
\]
This is the same as saying the remainder when $x$ is divided by 26
is the same as the remainder when $y$ is divided by 26.

If two numbers differ by a multiple of 26, we say that they are
\defone{congruent} mod 26.

Note that
\begin{align*}
26 &\equiv 0 \pmod{26} \\   
27 &\equiv 1 \pmod{26} \\
  28 &\equiv 2 \pmod{26} \\
  ...
\end{align*}
and
\begin{align*}
-1 &\equiv 25 \pmod{26} \\   
-2 &\equiv 24 \pmod{26} \\
-3 &\equiv 23 \pmod{26} \\
  ...
\end{align*}
So in the mod 26 world, since you are ignoring multiples of 26,
\textit{in some sense} there are only 26 numbers:
\[
  0, 1, 2, 3, ..., 25
\]
I'll write $\Z/26$ for this world of 26 values.
Remember that in this world, you can write the symbol
\[
  28
\]
but this is the same as 2 in $\Z/26$:
\[
  28 \equiv 2 \pmod{26}
\]
Of course in $\Z$, these symbols, i.e. $28$ and $2$, are different.

Most of the algebraic rules involving $+, -, *, 0, 1$ applies when
working with integers mod 26.
For instance suppose 
\[
  x \equiv y \pmod{26}
\]
where $x$ and $y$ are integers
(i.e., $x$ differs from $y$ by a multiple of 26), then
\[
  x + z \equiv y + z\pmod{26}
\]
where $z$ is an integer.
Likewise
from
\[
  x \equiv y \pmod{26}
\]
we get
\[
  x z \equiv y z \pmod{26}
\]
It's also true that
\[
  0 + x \equiv x \pmod{26}
\]
and 
\[
  1 \cdot x \equiv x \pmod{26}
\]

To be more precise, $\Z/26$ is a commutative ring.
It's a finite commutative ring with 26 values.
Let me rewrite the axioms for a commutative ring for $\Z/26$.

For $+$ on $\Z/26$, the properties are:
\begin{myenum}
\item $(r + s) \pmod{26}$ is in $\Z/26$ for all $r,s$ in $\Z/26$
\item 
  $(r + s) + t
  \equiv
  r + (s + t) \pmod{26}$
for all $r,s,t$ in $\Z/26$
\item
$r + 0 \equiv r \equiv 0 + r \pmod{26}$ 
for all
$r$ in $\Z/26$.
In fact $r'$ is just $(26 - r) \pmod{26}$.
For instance for $r = 2 \pmod{26}$, $r' = 26 - 2 = 24 \pmod{26}$.
\item For 
all $r$ in $\Z/26$ there is something in $\Z/26$ which we will call $r'$
such that $r + r' \equiv 0 \equiv r' \oplus r \pmod{26}$
\end{myenum}

For $\cdot$ the properties are:
\begin{myenum}
\item $r \cdot s \pmod{26}$ is in $\Z/26$ for all $r, s$ in $\Z/26$
\item $(r \cdot s) \cdot t \equiv r \cdot (s \cdot t) \pmod{26}$ for all $r,s,t$ in $\Z/26$
\item $r \cdot 1 \equiv r \equiv 1 \cdot r \pmod{26}$ for all $r$ in $\Z/26$
\item $r \cdot s \equiv s \cdot r \pmod{26}$ for all $r,s$ in $\Z/26$ 
\end{myenum}
The property involving both $+$ and $\cdot$ is
\begin{myenum}
\item 
$r \cdot (s + t) \equiv$
$r \cdot s + r \cdot t \pmod{26}$
\end{myenum}

It's also not too surprising that you can talk about mod $n$ for any
positive integer $n$.

If I don't say so, when I want you write some $x \pmod{26}$, I mean the $x$
such that $0 \leq x < 26$.
Of course there's no difference in mod 26 between 2 and 28.
But in mod 26, the values in $[0, 26)$ is the preferred range.
Also, when I say simplify $27 \pmod{26}$, I mean $1 \pmod{26}$.

Note that since $\Z/26$ is \textit{finite}, you can always solve
any equation in mod 26.
This is similar to boolean values: there are only two.
To solve a boolean equation, you just try all possible boolean values.
So to solve a $\Z/26$ equation, you just try all the 26 possible
values on all the variables that appear in the equation.

\input{exercises/exercise-0/main.tex}
\input{exercises/exercise-1/main.tex}
\input{exercises/exercise-2/main.tex}
\input{exercises/exercise-3/main.tex}
\input{exercises/exercise-4/main.tex}
\input{exercises/exercise-5/main.tex}
\input{exercises/exercise-6/main.tex}
\input{exercises/exercise-7/main.tex}
\input{exercises/exercise-8/main.tex}
\input{exercises/exercise-9/main.tex}
\input{exercises/exercise-10/main.tex}
\input{exercises/exercise-11/main.tex}
\input{exercises/exercise-12/main.tex}
\input{exercises/exercise-13/main.tex}
\input{exercises/exercise-14/main.tex}
\input{exercises/exercise-15/main.tex}
\input{exercises/exercise-16/main.tex}
\input{exercises/exercise-17/main.tex}
\input{exercises/exercise-18/main.tex}
\input{exercises/exercise-19/main.tex}
\input{exercises/exercise-20/main.tex}
\input{exercises/exercise-21/main.tex}
\input{exercises/exercise-22/main.tex}
\input{exercises/exercise-23/main.tex}
\input{exercises/exercise-24/main.tex}
\input{exercises/exercise-25/main.tex}
\input{exercises/exercise-26/main.tex}


