\sectionthree{Permutation cipher}
\begin{python0}
from solutions import *; clear()
\end{python0}

In the above ciphers, a character is replaced by another.
The
\defone{permutation cipher}
is different:
each character is \textit{moved} to a different
\textit{location} in the plaintext
to form the ciphertext.

Here's an example.
Look at this:
\[
  \pi =
  \begin{pmatrix}
    1 & 2 & 3 & 4 & 5 & 6 \\
    3 & 1 & 2 & 5 & 4 & 6 \\
  \end{pmatrix}
  \]
  It's just the function where a value in the top row maps to the
  corresponding value at the bottom row: 
  \begin{align*}
    \pi(1) &= 3 \\
    \pi(2) &= 1 \\
    \pi(3) &= 2 \\
    \vdots &\textwhite{=} \ \vdots \\
    \pi(6) &= 6 \\
  \end{align*}
Let $\pi$ is called a
permutation
of $\{1,2,3,4,5,6\}$.
A
\defone{permutation}
is simply a bijection (1--1 and onto) function
from
$\{1, 2, 3, ..., n\}$
to
$\{1, 2, 3, ..., n\}$.
As a shorthand, the above permutation
\[
  \pi =
  \begin{pmatrix}
    1 & 2 & 3 & 4 & 5 & 6 \\
    3 & 1 & 2 & 5 & 4 & 6 \\
  \end{pmatrix}
\]
is also written as
\[
  \pi = (1 \ \ 3 \ \ 2)(4 \ \ 5)(6)
\]
This is called the
\defone{cycle notation}
of $\pi$.
In the above example, we say that the permutation $\pi$ has
\defone{length}
6.

\input{exercises/permutation-0/main.tex}

How does a permutation give you an encryption?
For the permutation $\pi = (1 \ \ 3 \ \ 2)(4 \ \ 5)(6)$,
you do this:
\begin{enumerate}[nosep]
  \item
    character at position 1 goes to position 3,
  \item
    character at position 2 goes to position 1,
  \item etc.
\end{enumerate}
In the example above, the $n$ is 6.
For instance for the string \texttt{marvin}
and the permutation $\pi = (1 \ \ 3 \ \ 2)(4 \ \ 5)(6)$,
the since in the permutation $1 \mapsto 3$, the character at
position 1 goes to position 3:
\begin{smallconsole}
    1 2 3 4 5 6
    m a r v i n
        m
\end{smallconsole}
since $3 \mapsto 2$, the character at
position 3 goes to position 2:
\begin{smallconsole}
    1 2 3 4 5 6
    m a r v i n
      r m
\end{smallconsole}
Etc.
Altogether I get
\[
E(\pi, \texttt{marvin}) = \texttt{armivn}
\]
This assumes your string has a length which is a multiple of 6.
So you might need to add some dummy data at the end of your string.


Note that in the above example, you have to encrypt substrings of length 5.
Therefore if the original plaintext does not have a length that is a
multiple of 5, you would need to pad it with some characters until the
length is a multiple of 5.

\input{exercises/permutation-1/main.tex}

The decryption is the same algorithm as the encryption
except that the permutation is read in the \lq\lq opposite direction''.
If
\[
  \pi =
  \begin{pmatrix}
    1 & 2 & 3 & 4 & 5 \\
    3 & 1 & 2 & 5 & 4 \\
  \end{pmatrix}
\]
then the inverse of $\pi$, denoted by $\pi^{-1}$ is the \lq\lq opposite''
of the above:
\[
  \pi^{-1} =
  \begin{pmatrix}
    3 & 1 & 2 & 5 & 4 \\
    1 & 2 & 3 & 4 & 5 \\
  \end{pmatrix}
\]
(do you see the row switch?) and to make things look nice, you arrange the
top row:
\[
  \pi^{-1} =
  \begin{pmatrix}
    1 & 2 & 3 & 4 & 5 \\
    2 & 3 & 1 & 5 & 4 \\
  \end{pmatrix}
\]
Using the cycle notation,
\[
  \pi^{-1} = (1 \ \ 2 \ \ 3)(4 \ \ 5)
\]
So
\[
  D(\pi, y)
\]
is the same as
\[
  E(\pi^{-1}, y)
\]
So
\begin{align*}
  D((1 \ \ 3 \ \ 2)(4 \ \ 5), \texttt{armivn})
  &= E((1 \ \ 3 \ \ 2)(4 \ \ 5)^{-1}, \texttt{armivn}) \\
  &= E((1 \ \ 2 \ \ 3)(4 \ \ 5), \texttt{armivn}) \\
  &= \texttt{marvin}
\end{align*}

\input{exercises/permutation-2/main.tex}
\input{exercises/permutation-3/main.tex}
\input{exercises/permutation-4/main.tex}


\textsc{Breaking the permutation cipher}

Frequency analysis obviously won't help: each letter is not replaced by another.
It's the position of a letter that is changed.

Now suppose the permutation is
\[
  \pi =
  \begin{pmatrix}
    1 & 2 & 3 & 4 & 5 \\
    3 & 1 & 2 & 5 & 4 \\
  \end{pmatrix}
\]
Suppose you know the length of the permutation is 5.
After breaking your ciphertext into substrings of length 5,
suppose one of these substrings is
\begin{Verbatim}
  hatoe
\end{Verbatim}
We know that \verb!th! is a commonly occurring digram.
Then it's natural to suspect that this comes from the encryption of 5 letters containing
\verb!th!,
either
\begin{Verbatim}
  plaintext : ..... th... ...
  ciphertext: ..... hatoe ...
\end{Verbatim}
or
\begin{Verbatim}
  plaintext : ..... .th.. ...
  ciphertext: ..... hatoe ...
\end{Verbatim}
or
\begin{Verbatim}
  plaintext : ..... ..th. ...
  ciphertext: ..... hatoe ...
\end{Verbatim}
or
\begin{Verbatim}
  plaintext : ..... ...th ...
  ciphertext: ..... hatoe ...
\end{Verbatim}
Whereas for a substring of length 5, there are $5! = 120$ possible
permutations,
once two is fixed, there are $3! = 6$ possible permutations.
This might cut down on the search for the key and help break the ciphertext.
If \verb!th! is not found or the above analysis involving \verb!th!
does not work, you go on to the next commonly occurring digram.

Of course you do not know if the length of the permutation is length 5.
You therefore have to do a loop over all possible permutation lengths.
Of course the key length cannot be 1 since that would be doing nothing!
For key of size two, the permutation must be $(1, 2)$.
For a key of size three, there are $3! = 6$ permutations:
There are 6 possible permutations:
$(1)(2)(3)$, $(1)(2, 3)$, $(2)(1, 3)$, $(3)(1, 2)$,
$(1, 2, 3)$, $(1, 3, 2)$.
Since there are not many permutations of length 3, for this case,
you might want to simply try all permutations.
In general, you try key lengths dividing the length of the ciphertext.

Note that it's also possible that the \verb!th! in the plaintext
is broken up with the \verb!t! is one substring of length 5
and \verb!h! in the following substring.
For instance, here is a case:
\begin{Verbatim}[commandchars=\\\{\}]
  plaintext : cant\underline{h} \underline{e}fish         
  ciphertext: acn\underline{h}t f\underline{e}ihs
\end{Verbatim}
Of course it's also OK if you do not wish to consider cases where the
digram is split across two substrings, since in the above case,
you would have picked up the \verb!he! when you consider a key of length (say) 10.

By looking at lots of digrams and trigrams, if the ciphertext is long enough, you
should be able to break the code.

\newpage
\begin{eg}
  Let me break the following permutation ciphertext
\[
  \texttt{imahsnrwnesotanfusefvitieedoslwswma}
\]
\end{eg}

\SOLUTION

The length is 35. Since the length of the permutation must divide 35,
it must be 1, 5, 7, 35.


\textsc{Length 1.}
Of course if the length is 1, then the ciphertext is the same as the
plaintext. But the plaintext is meaningless. Therefore the length
cannot be 1.

\textsc{Length 5.}
Assume the key length is 5.
I break up the string
above into substrings of length 5 to make it easier to read:
\begin{Verbatim}
  imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}


\textsc{Length 5. Digram \texttt{th}.}
I don't see any \verb!th! in each substrings or consecutive substrings.


\textsc{Length 5. Digram \texttt{he}.}
I do see \verb!he!:
\begin{Verbatim}[commandchars=\\\{\}]
  ima\underline{h}s nrwn\underline{e} sotan fusef vitie edosl wswma
\end{Verbatim}
If you assume these two characters come from \verb!he!, then it has to come from this:
\begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  ....h e.... ..... ..... ..... ..... .....
 ciphertext: ima\underline{h}s nrwn\underline{e} sotan fusef vitie edosl wswma
\end{Verbatim}
which means the permutation contains $1 \mapsto 5, 5 \mapsto 4$.
If that's the case, we have
\begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  s...h e...n n...a f...e e...i l...s a...m
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}
Look at just the first group of 5 characters
the \verb!s...h! can only be
\verb!simah!,
\verb!siamh!,
\verb!saimh!,
\verb!samih!,
\verb!smiah!, or
\verb!smaih!.
The only promising one is \verb!siamh! (maybe \lq\lq Siam has cats ...?").
Using this we get
\begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  siamh enwrn nstoa ffsue evtii leods awwsm
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}
which is meaningless.

\newpage
[force newpage]

\textsc{Length 5. Digram \texttt{in}.}
The next digram to try is \verb!in!:
\begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  ....i n.... ..... ..... ..... ..... .....
 ciphertext: \underline{i}mahs \underline{n}rwne sotan fusef vitie edosl wswma
\end{Verbatim}
or
\begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  ....i n.... ..... ..... ..... ..... .....
 ciphertext: \underline{i}mahs nrw\underline{n}e sotan fusef vitie edosl wswma
\end{Verbatim}
The first is not possible. (Why?)
% the i means 5->1 and the n means 1->1. cannot have 5->1,1->1 for permutation.
The second gives us
\begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  h...i n...n a...s e...f i...v s...e m...w
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}
Looking at just the first 5 characters,
the only possible decryptions are
\verb!hmasi!,
\verb!hmsai!,
\verb!hamsi!,
\verb!hasmi!,
\verb!hsami!,
\verb!hsmai!.
Only
\verb!hamsi!,
\verb!hasmi!,
are promising.
From \verb!hamsi!,
\begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  hamsi nwren atons esuff itiev sodle mwsaw
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}
which is meaningless.
From \verb!hasmi!,
\begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  hasmi nwern atnos esfuf iteiv solde mwasw
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}
which is meaningless.

\textsc{Length 5. Digram \texttt{er}.}
The next digram to try is \verb!er!:
\begin{itemize}
  \li From
  \begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  ..... er... ..... ..... ..... ..... .....
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}
we get
  \begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  sm... er... no... fu... ei... ld... as...
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}
The first 5 characters can only be
\verb!smahs!,
\verb!smash!,
\verb!smhas!,
\verb!smhsa!,
\verb!smsah!,
\verb!smsha!. Only \verb!smash! seems to form a word.
(\verb!smhas! might be for instance \texttt{a\ cha\underline{sm\ has}\ separated\ us},
but the \verb!smhas! is the \textit{beginning} of the plaintext.)
Using that, we get
  \begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  smash erwen notna fusfe eitei ldols aswam
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}
which is not meaningful.

  \li From
  \begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  ..... .er.. ..... ..... ..... ..... .....
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}
we get
  \begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  .sm.. .er.. .no.. .fu.. .ei.. .ld.. .as..
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}
The first 5 characters can only be
\verb!ismah!,
\verb!ismha!,
\verb!asmih!,
\verb!asmhi!,
\verb!hsmia!,
\verb!hsmai!,
none of which is meaningful,
except possibly for
\verb!ismah! (example: \texttt{\underline{is\ mah}i\ mahi\ a\ fish?}),
\verb!ismha! (example: \texttt{a sch\underline{ism\ ha}ppened overnight}),
\verb!asmih! (example: \texttt{protopl\underline{asm i h}ave}),
\verb!asmhi! (example: \texttt{protopl\underline{asm h}ave i})
but you can check that these do not lead
to anything meaningful.


\li From
  \begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  ..... ..er. ..... ..... ..... ..... .....
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}
we get
  \begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  ..sm. ..er. ..no. ..fu. ..ei. ..ld. ..as.
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}
The first 5 characters can only be
\verb!iasmh!,
\verb!ihsma!,
\verb!aismh!,
\verb!ahsmi!,
\verb!hisma!,
\verb!hasmi!.
Only
\verb!ahsmi!,
\verb!hisma!
and
\verb!hasmi!
seem promising.
For \verb!ahsmi!, we get
  \begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  ahsmi wnern tanos sefuf tieiv oslde wmasw
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
\end{Verbatim}
which is meaningless.
For \verb!hisma!, we get
  \begin{Verbatim}[commandchars=\\\{\}]
 plaintext:  hisma nnerw asnot effus iveit seldo mwasw
 ciphertext: imahs nrwne sotan fusef vitie edosl wswma
 \end{Verbatim}
 which finally gives us the plaintext
 \[
   \texttt{hismannerwasnoteffusiveitseldomwasw}
 \]
 This gives us
 \[
    \texttt{His manner was not effusive.\ It seldom was.}
 \]
 % hismannerwasnoteffusiveitseldomwasw
 The last \verb!w! is redundant (it's a padding).
 \end{itemize}

 The key (i.e., permutation) is
 $(1,4,2)(3,5)$ with a key length of 5.
 \qed

\begin{comment}
plainttext: His manner was not effusive. It seldom was.
key = (1, 4, 2)(3, 5)

hisma - imahs
nnerw - nrwne
asnot - sotan
effus - fusef
iveit - vitie
seldo - edosl
mwasw - wswma
\end{comment}

\input{exercises/permutation-5/main.tex}
\input{exercises/permutation-6/main.tex}
\input{exercises/permutation-7/main.tex}
\input{exercises/permutation-8/main.tex}
\input{exercises/permutation-9/main.tex}
\input{exercises/permutation-10/main.tex}


