\sectionthree{Hill cipher}
\begin{python0}
from solutions import *; clear()
\end{python0}

\defone{Hill's cipher}
was announced in 1929.
This cipher works with a block of $n$ characters at a time.
For both encryption and decryption, you 
treat each block of $n$ characters (or mod 26 integers) as a column
vector
and we multiply it with an $n$--by--$n$ matrix.
Of course everything is done in mod 26.
Here are more details.

Recall how to multiple a $n$--by--$n$ matrix with a column vector of
size $n$.
I'll do this for $n = 2$.
\[
\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}
\begin{pmatrix}
  x \\
  y 
\end{pmatrix}
=
\begin{pmatrix}
  ax + by \\
  cx + dy 
\end{pmatrix}
\]
(If you have not seen this before take the computer graphics class CISS380.
You can't call yourself a serious computer scienctist if you can't do matrix
multiplication.
Most top CS programs actually has a linear algebra requirement.) 
That's matrix multiplication where all numbers are in $\R$.
The (multiplicative)
\defone{inverse matrix}
of
\[
\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}
\]
is
given by
\[
\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}^{-1}
=
\frac{1}{ad - bc}
\begin{pmatrix}
  d & -b \\
  -c & a 
\end{pmatrix}
\]
The expression $ad - bc$ is called the
\defone{determinant}
of
$\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}$ and is written
\[
\det
\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}
= ad - bc
\]
If a matrix has an inverse, we say that the matrix is
\defone{invertible}.

Not only can you multiply a 2-by-2 matrix
with a 2-by-1, you can actually also multiply
a 2-by-2 matrix
with a 2-by-2.
Here's how you do it:
\[
\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}
\begin{pmatrix}
  A & C \\
  B & D 
\end{pmatrix}
=
\begin{pmatrix}
  aA + bB &   aC + bD \\
  cA + dB &   cC + dD \\
\end{pmatrix}
\]


\input{exercises/hill-0/main.tex}

Define the following:
  \[
  I =
  \begin{pmatrix}
    1 & 0 \\
    0 & 1 
  \end{pmatrix}
  \]
  This is called the
  (2-by-2)
  \defone{identity matrix}.


\input{exercises/hill-1/main.tex}
\input{exercises/hill-2/main.tex}

The set of $n$--by--$n$ matrices with $\R$ values is usually denoted by
$\operatorname{M}_{n}(\R)$.
The set of invertible matrices is usually denoted by
$\operatorname{GL}_{n}(\R)$. This is called the
\defone{general linear group} of $n$-by-$n$ matrices over $\R$.

It turns out that everything works just as fine if the numbers are
in mod 26.
You just need to be careful with the determinant of the inverse
computation.
Recall that
\[
\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}^{-1}
=
\frac{1}{ad - bc}
\begin{pmatrix}
  d & -b \\
  -c & a 
\end{pmatrix}
\]
for values in $\R$.
This is the same for values in $\Z/26$ as long as you think of
\[
\frac{1}{ad - bc}
\]
as the multiplicative inverse of $ad - bd$ in $\Z/26$.

\input{exercises/hill-3/main.tex}
\input{exercises/hill-4/main.tex}

This should not be surprising:
The set of $n$--by--$n$ matrices with mod 26 values is usually denoted by
$\operatorname{M}_{n}(\Z/26)$.
The set of invertible matrices is usually denoted by
$\operatorname{GL}_{n}(\Z/26)$.

\input{exercises/hill-5/main.tex}

Note that in the above, since the block size is 2, you would need to
encrypt a string with a string length that is a multiple of $2$.
If it's not, you can (for instance) pad with some character.

The important thing to observe is that if
\[
  E(M, (x, y)) = (x', y')
  \]
  ($M$ is a matrix key)
note that $x$ takes part in the value $x'$ \textit{and} $y'$.
Same thing for $y'$.
This means that if I change one character in my plaintext, \textit{two}
character in the ciphertext is changed.
Informally this means that the statistical behavior of $x$
is spread out across multiple characters.
Note only that $x'$ also depends on several values of the key (the matrix).
This makes it much hard to analyze the key one part at a time when
studying ciphertexts.
This makes it a lot harder to break Hill's cipher, especially when
the matrice size is larger, say the matrix is 32-by-32.

The above two important observations
were later formalized and studied by Claude Shannon in 1945.
\begin{enumerate}[nosep]
\item The fact that a character of the ciphertext
depends on several parts of the key is the
concept of
\defone{confusion}
and
\item the fact that one character in the plaintext
influences multiple characters in the ciphertext
is called the concept of
\defone{diffusion}
\end{enumerate}
More on that in a much later chapter.
\input{exercises/hill-6/main.tex}
\input{exercises/hill-7/main.tex}
\input{exercises/hill-8/main.tex}
\input{exercises/hill-9/main.tex}
\input{exercises/hill-10/main.tex}

Collecting up all the information above,
formally, the Hill's cipher of block size $n$ (just think of $n = 2$)
is defined as follows:
Let
\[
P = C = \{0,...,25\}^n = (\Z/26)^n
\]
and
\[
K = \operatorname{GL}_n(\Z/26)
\]
The encryption function $E$
\[
E : K \times P \rightarrow C
\]
is defined to be
\[
E(M, x) = Mx
\]
and
\[
D : K \times C \rightarrow P
\]
is defined to be
\[
D(M, x) = M^{-1}x
\]
And $(E,D)$ is a cipher because
\[
D(M, E(M, x)) = D(M, Mx) = M^{-1}Mx = Ix = x
\]

\input{exercises/hill-11/main.tex}

\newpage
\textsc{Breaking Hill's cipher}

Now how do you break the Hill's cipher?

Let's stick to a 2-by-2 key.
Suppose you have the following:
you have the two plaintexts
$x_1 = (A, B)$ and $x_2 = (C, D)$
and their encryptions
$y_1 = (A', B')$ and $y_2 = (C', D')$.
Therefore
\[
\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}
\begin{pmatrix}
  A \\
  B 
\end{pmatrix}
=
\begin{pmatrix}
  aA + bB \\
  cA + dB 
\end{pmatrix}
=
\begin{pmatrix}
  A' \\
  B' 
\end{pmatrix}
\]
and
\[
\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}
\begin{pmatrix}
  C \\
  D 
\end{pmatrix}
=
\begin{pmatrix}
  aC + bD \\
  cC + dD 
\end{pmatrix}
=
\begin{pmatrix}
  A' \\
  B' 
\end{pmatrix}
\]
In the above, the unknowns are $a,b,c,d$, i.e., the key.
These two equations can be combined to get an equation involving
the product of 2-by-2 matrices:
\[
\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}
\begin{pmatrix}
  A & C \\
  B & D 
\end{pmatrix}
=
\begin{pmatrix}
  aA + bB & aC + bD \\
  cA + dB & cC + dD \\
\end{pmatrix}
=
\begin{pmatrix}
  A' & C' \\
  B' & D'
\end{pmatrix}
\]
You now have this matrix equation
\[
\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}
\begin{pmatrix}
  A & C \\
  B & D 
\end{pmatrix}
=
\begin{pmatrix}
  A' & C' \\
  B' & D'
\end{pmatrix}
\]
Note that you have the values of $A,B,C,D,A',B',C',D'$.
The unknowns are $a,b,c,d$.
What do you do?
You multiply on the right by the inverse matrix of
$\begin{pmatrix}
  A & C \\
  B & D 
\end{pmatrix}
$
to get
\[
\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}
\begin{pmatrix}
  A & C \\
  B & D 
\end{pmatrix}
\begin{pmatrix}
  A & C \\
  B & D 
\end{pmatrix}^{-1}
=
\begin{pmatrix}
  A' & C' \\
  B' & D'
\end{pmatrix}
\begin{pmatrix}
  A & C \\
  B & D 
\end{pmatrix}^{-1}
\]
which gives you
\[
\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}
I
=
\begin{pmatrix}
  A' & C' \\
  B' & D'
\end{pmatrix}
\begin{pmatrix}
  A & C \\
  B & D 
\end{pmatrix}^{-1}
\]
which finally gives you the key
\[
\begin{pmatrix}
  a & b \\
  c & d 
\end{pmatrix}
=
\begin{pmatrix}
  A' & C' \\
  B' & D'
\end{pmatrix}
\begin{pmatrix}
  A & C \\
  B & D 
\end{pmatrix}^{-1}
\]
There's an assumption above:
we have to assume that
$
\begin{pmatrix}
  A & C \\
  B & D 
\end{pmatrix}
$
is invertible.
It's possible the first 4 characters of the original message
does not form an invertible matrix.
But if you have enough pairs of plaintext and corresponding ciphertext blocks,
you hope to be able to get an invertible matrix and perform
the above computation.
Therefore we are assuming the known plaintext attack model.
Or we are assuming the chosen plaintext attack model.
