\tinysidebar{\debug{exercises/{exercise-13/question.tex}}}
Consider
  \[
    5x^2 - 8y^2 = 3 z^2
  \]
  where we are only interested in finding integer solutions.
  Of course $(0,0,0)$ is a solution, but that's easy. (Right?)
  In general, an equation like
  \[
    ax^p + b y^q = c z^r
  \]
  where $p,q,r > 0$ always has $(0,0,0)$ as a solution.
  $(0,0,0)$ is sometimes called the trivial solution.
  So we might as well assume $(x,y,z) \neq (0,0,0)$, i.e.,
  assume they are not all zero.
  By the way the polynomial
  \[
    5x^2 - 8y^2 - 3 z^2    
  \]
  is a sum of terms where the number of variables appearing in each term is the same, i.e., 2.
  Such polynomial are is said to be homogeneous of degree 2.
  

  How many solutions can you find?
  Is there any at all?
  Is there a finite number of solutions?
  If there are infinitely many solutions, can you write them down?
  If not all of them, maybe an infinite family of them.
  For instance it would be nice to say:
  \lq\lq For any integer $n$, $x=n, y = 2n, z = n + 5$ is a solution''.
  This would be a 1--parameter family of solutions.
  
  \begin{myenum} 

  \item What's the first thing you should do?
    % Write a program to find solution by brute force in [0,50] x [0,50] x [0,50].

  \item Prove that if $x,y,z$ is a solution, then
    $nx,ny,nz$ is also a solution for any integer $n$.
    
  \item
    In case you can't see a solution right away, do the following.
    If $(x,y,z)$ are solutions,
    show that $x$ and $z$ must be even.
    (HINT: mod 4.)
    This is helpful since we won't have to check the odd $x$ or odd $z$ caaes.
    That cuts down a brute force search down by 3/4.
    
  \item
    Continuing the above:
    let $x = 2a$, $y = b$, and $z = 2c$, substitute,
    and you'll get a new equation in $a,b,c$.

  \item
    The equation in $a,b,c$ is easier than the one in $x,y,z$.
    Why? Because a solution in the equation in $a,b,c$ for the above question
    would
    correspond to a solution $x,y,z$ in the original equatison and $a,b,c$
    solution is smaller. 
    Can you now find some solutions for the equation in $a,b,c$?

  \item
    I won't go further. But for those who want to solve this
    probably complete, let me just say that once you have one
    one solution, you can parametrize all solutions using
    two integer parameters.
  \end{myenum}
